Bifurcation or state tipping: assessing transition type in a model trophic cascade

### 2018-12-11 paper/paper.Rmd
WIP

Central to resolving this question is untangling a second issue that frequently arises in the discussion of alternative stable state dynamics in ecological systems: the issue of dimension.  The S-tipping and B-tipping transitions illustrated in Figure \ref{fig1} are fundamentally rooted in a one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e. population density of adult bass).  Ecosystems are notoriously more complex, and this lake ecosystem is no exception. Here, we analyze a previously published model of a trophic cascade to determine whether manipulating the top predator constitutes B-tipping or S-tipping. Common versions of this model include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}). We perform dimension reduction on the full 5D model. This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.

@@

This analysis also highlights the importance of what might appear to be a merely semantic issue that in fact corresponds to fundamentally different dynamics. In this formulation, the control parameter $q$ corresponds to harvest (positive $q$) or stocking (negative $q$) of adult bass in the lake. It is this aspect that first gives the appearance of state tipping dynamics, since it suggests the state is being somehow "directly manipulated". Yet from a mathematical standpoint, it is irrelevant if the control parameter corresponds to some environmental or policy variable (fishing regulations, fish migration into the lake), which in turn increases or decreases net mortality proportionally, or if the manipulation directly adds or removes adult bass -- the feature that distinguishes this manipulation from a S-tipping transition comes down to the timescales involved. Slowly increasing in the amount of bass stocked in the lake (i.e. decreasing $q$, effectively running the scenario of increasing fishing harvests in reverse) gives the rest of the food web time to adjust.  In contrast, a single large introduction of many bass would create an S-tipping transition, driving the system immediately into an alternate state without exhibiting CSD phenomena needed to evaluate early warning signals of a tipping point.

>>>>

Central to resolving this question is untangling a second issue that frequently arises in the discussion of alternative stable state dynamics in ecological systems: the issue of dimension.   Ecosystems are notoriously complex, and this lake ecosystem is no exception. Here, we analyze a previously published model of a trophic cascade to determine whether manipulating the top predator constitutes B-tipping or S-tipping. Common versions of this model include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}). In sharp contast to this complexity, both intutition and theory of critical transitions have been centered on one-dimensional intpretations. The S-tipping and B-tipping transitions illustrated in Figure \ref{fig1} are fundamentally rooted in a one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e. population density of adult bass).  Similarly, Ball-in-cup style depictions of alternative stable states and sudden transitions present only a one-dimensional dipiction of a system state. While it may be tempting to dismiss such simple depictions as out-of-hand as mere cartoons, it is in fact possible to map the full five-dimensional model explicitly into a one-dimensional approximation under a few simple assumptions.  Here, we present the full dimension reduction on the full 5D model analytically, and compare this approximation to numerical simulations of the original system. This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.

@@

This analysis also highlights the importance of what might appear to be a merely semantic issue that in fact corresponds to fundamentally different dynamics. In this formulation, the control parameter $q$ corresponds to harvest (positive $q$) or stocking (negative $q$) of adult bass in the lake. It is this aspect that first gives the appearance of state tipping dynamics, since it suggests the state is being somehow "directly manipulated". Yet from a mathematical standpoint, it is irrelevant if the control parameter corresponds to some environmental or policy variable (fishing regulations, fish migration into the lake), which in turn increases or decreases net mortality proportionally, or if the manipulation directly adds or removes adult bass -- the feature that distinguishes this manipulation from a S-tipping transition comes down to the scales involved. This can be seen most clearly though an analogy of considering the same model applied to a larger scale, such as marine fisheries.  In fisheries models, one would routinely examine the dynamics of the fish population biomass $B$ with respect to different possibly parameters for harvests or fishing effort.  Even though harvest may be measured in the same units as the total biomass, it is natural to think of the two quantites very differently.  A manager can adjust the harvest effort by changing quotas, limiting permits for fishing boats etc.  Slowly increasing in the amount of bass stocked in the lake (i.e. decreasing $q$, effectively running the scenario of increasing fishing harvests in reverse) gives the rest of the food web time to adjust.  In contrast, a single large introduction of many bass would create an S-tipping transition, driving the system immediately into an alternate state without exhibiting CSD phenomena needed to evaluate early warning signals of a tipping point.
